Confluence in Lens Synthesis

### 2020-06-11 papers/BijectiveLenses/IWC/main.tex
update iwc

  Our proof that we can synthesize all string lenses involves proving a
  confluence-like property, indexed by a an additional binary relation $R$. We
  call this property $R$-confluence. In
  this model, standard confluence is the specific case where $R$ is equality. We
  show how our proof that the transitive closure of a rewrite system is
  $R$-confluent differs from the proof strategy for standard confluence.
\end{abstract}

@@

Optician~\cite{optician}, an extension of Boomerang, makes bidirectional
programming easier by supporting synthesis of bidirectional string
transformations. More specifically, it takes as input two regular expressions
($\Regex$ and $\RegexAlt$, which serve as the type of a Boomerang lens) and a set of examples specifying
input-output behavior, and synthesizes a well-typed lens between the languages
of those regular expressions.

To indicate a lens $\Lens$ is well typed, and converts between the languages of
$\Regex$ and $\RegexAlt$, we write $\Lens : \Regex \Leftrightarrow \RegexAlt$
For synthesis, given $\Regex$ and $\RegexAlt$, we must find a lens $\Lens$ such
that $\Lens : \Regex \Leftrightarrow \RegexAlt$. In the context of lens
synthesis, there are three sorts of lens typing rules: syntax-directed rules,
composition, and type equivalence.

@@

In our search algorithm, we only search through equivalent regular expressions
once, before processing any syntax directed-rules. We formalize this in the
typing of DNF regular expressions by only permitting the application of
type-equivalence once, after all syntactic rules have been applied. This is
enforced by having two typing judgements: one for the ``rewriteless'' type of
the lens (meaning no type-equivalence rules were applied) and one of the
``full'' type of the lens. If $\DNFLens \OfRewritelessType \DNFRegex
\Leftrightarrow \DNFRegexAlt$, then $\DNFLens$ is a DNF lens of rewriteless type
$\DNFRegex \Leftrightarrow \DNFRegexAlt$. If $\DNFLens \OfType \DNFRegex
\Leftrightarrow \DNFRegexAlt$, then $\DNFLens$ is a DNF lens of full type
$\DNFRegex \Leftrightarrow \DNFRegexAlt$. The following rule is used to get the
full type of a DNF lens from the rewriteless type.
\begin{mathpar}
  \inferrule[\RewriteDNFRegexLensRule{}]
  {
    \DNFRegex' \StarOf{\Rewrite} \DNFRegex\\
    \DNFRegexAlt' \StarOf{\Rewrite} \DNFRegexAlt \\
    \DNFLens \OfRewritelessType \MapsBetweenTypeOf{\DNFRegex}{\DNFRegexAlt}
  }
  {
    \DNFLens \OfType \MapsBetweenTypeOf{\DNFRegex'}{\DNFRegexAlt'}
  }
\end{mathpar}
Note that instead of using directionless equivalences,
\RewriteDNFRegexLensRule{} uses directioned rewrites. The relationship between
the two is formalized by the following theorem:
\begin{theorem}
  \label{thm:thm1}
  If $\Regex \SSREquiv \RegexAlt$, then $\ToDNFRegexOf{\Regex}
  \equiv_{\rightarrow} \ToDNFRegexOf{\RegexAlt}$, where $\ToDNFRegexOf{\Regex}$
  and $\ToDNFRegexOf{\RegexAlt}$ are $\Regex$ and $\RegexAlt$ in DNF form
  (respectively), and $\equiv_{\rightarrow}$ is the transitive, reflexive, and
  symmetric closure of $\rightarrow$.
\end{theorem}

Proving that our search procedure can generate any lens reduces to proving DNF
lenses complete with respect to our standard lens language. In particular, we
wish the prove the following theorem:
\begin{theorem}
  If $\Lens : \Regex \Leftrightarrow \RegexAlt$, then there exists a DNF lens,
  $\DNFLens$, such that $\DNFLens \OfType \ToDNFRegexOf{\Regex} \Leftrightarrow
  \ToDNFRegexOf{\RegexAlt}$ and the semantics of $\Lens$ and $\DNFLens$
  are equivalent.
\end{theorem}

We prove this propety by induction on the structure of the typing derivation.
The particular difficulty lies in the lens equivalence rule. We begin this case
below:

\begin{mathpar}
  \inferrule
  {
    \Lens \OfType \Regex_1 \Leftrightarrow \RegexAlt_1\\
    \Regex \SSREquiv \Regex_2\\
    \RegexAlt \SSREquiv \RegexAlt_2
  }
  {
    \Lens \OfType \Regex_2 \Leftrightarrow \RegexAlt_2
  }
\end{mathpar}

@@

We first show that there exists $\DNFRegex_2$ and $\DNFRegexAlt_2$ such that
$\ToDNFRegexOf{\Regex_2} \StarOf{\rightarrow} \DNFRegex_2$ and
$\ToDNFRegexOf{\DNFRegex_1} \StarOf{\rightarrow} \DNFRegex_2$ and
$\ToDNFRegexOf{\RegexAlt_2} \StarOf{\rightarrow} \DNFRegexAlt_2$ and
$\ToDNFRegexOf{\DNFRegexAlt_1} \StarOf{\rightarrow} \DNFRegexAlt_2$. To do this,
we first prove that $\StarOf{\rightarrow}$ is confluent. By
Theorem~\ref{thm:thm1}, we know that $\ToDNFRegexOf{\Regex_1} \equiv_\rightarrow
\ToDNFRegexOf{\Regex_2}$. Such $\DNFRegex_2$ and $\DNFRegexAlt_2$ are guaranteed
to exist by a corollary to confluence.

\begin{figure}
  \centering
    \includegraphics[scale=.4]{equiv-case.pdf}
    \caption{Diagram constructing a well-typed DNF lens between
      \ToDNFRegexOf{\Regex_2} and \ToDNFRegexOf{\RegexAlt_2}.}
    \label{fig:equiv-case}
\end{figure}

After this, we have $\DNFLens \OfRewritelessType \DNFRegex_1 \Leftrightarrow
\DNFRegexAlt_1$ and $\DNFRegex_1 \StarOf{\rightarrow} \DNFRegex_2$ and
$\DNFRegexAlt_1 \StarOf{\rightarrow} \DNFRegexAlt_2$. If we can prove a
confluence-like property that would show the existance of some $\DNFLens'
\OfRewritelessType \DNFRegex_3 \Leftrightarrow \DNFRegexAlt_3$, where
$\DNFRegex_2 \StarOf{\rightarrow} \DNFRegex_3$ and $\DNFRegexAlt_2
\StarOf{\rightarrow} \DNFRegexAlt_3$, we would be done. This property is
$R$-confluence for a properly chosen $R$. This case is diagrammed in
Figure~\ref{fig:equiv-case}.

\section{$R$-Confluence Formulation}
Let $S$ be an underlying set, and $\rightarrow$ and $R$ be binary relations. The
indexed rewrite system $(S,\rightarrow)$ is $R$-confluent, if for all $s_1,s_2 \in
S$, if $R(s_1,s_2)$, $s_1 \rightarrow s_1'$, and $s_2 \rightarrow s_2'$, then
there exist $s_1''$ and $s_2''$ such that $R(s_1'',s_2'')$, $s_1' \rightarrow
s_1''$, and $s_2 \rightarrow s_2''$.

@@

If we followed the standard way to prove a rewrite system confluent, we would
find a set of rewrite rules $\ParallelRewrite$ such that
$\StarOf{\ParallelRewrite} = \StarOf{\Rewrite}$ and $(S,\ParallelRewrite)$ is
$R_\DNFLens$-confluent. Then, we need to prove that
$(S,\StarOf{\ParallelRewrite})$ is $R_\DNFLens$-confluent from knowledge that
$(S,\ParallelRewrite)$ is $R_\DNFLens$-confluent. However, we will show this
approach does not work.

@@

\begin{figure}
  \centering
    \includegraphics[scale=.4]{proof-attempt.pdf}
    \caption{If trying to prove $R$-confluence for the transitive closure of a
      rewrite system in the standard way, we get stuck at this point in the
      inductive case.}
    \label{fig:proof-attempt}
\end{figure}

The standard way of proving $(S,\StarOf{\rightarrow})$ to be $=$-confluent is to
first prove $(S,\rightarrow)$ to be $=$-confluent, then to perform two
inductions, first on the left arrows then on the right arrows. This approach
does not work for arbitrary $R$-confluence. Consider the inductive case, where
$R(s_1,t_1)$ and $s_1 \StarOf{\rightarrow} s_2$ and $s_2 \rightarrow s_3$ and
$t_1 \StarOf{\rightarrow} t_2$, shown in Figure~\ref{fig:proof-attempt}. By
applying the induction hypothesis, we know there exists $s_4, t_4$ such that $s_2
\rightarrow s_4$ and $t_2 \StarOf{\rightarrow} t_4$, and $R(s_4,t_4)$. We know
that $s_2 \rightarrow s_3$ and $s_2 \rightarrow s_4$, but we only know that
$(S,\rightarrow)$ is $R$-confluent, and so we are now stuck.

To get around this issue, we require a different set of properties.
\begin{enumerate}
\item $(S,\rightarrow)$ must be $=$-confluent.
\item $(S,\rightarrow)$ must be \emph{$R$-bisimilar}. In other words,
  if $R(s_1,t_1)$, and $s_1 \rightarrow s_2$, then there exists $t_2$ such that
  $t_1 \rightarrow t_2$ and $R(s_2,t_2)$; and if $R(s_1,t_1)$, and $t_1
  \rightarrow t_2$, then there exists $s_2$ such that $s_1 \rightarrow s_2$ and
  $R(s_2,t_2)$.
\end{enumerate}

Now we can proceed with our proof.

\begin{theorem}
  \label{lem:right-side}
  Let $(S,\rightarrow)$ be $=$-confluent, and $(S,\rightarrow)$ be
  $R$-bisimilar. If $R(s_1,t_1)$ and $s_1 \StarOf{\rightarrow} s_2$ and $t_1
  \StarOf{\rightarrow} t_2$, then there exists $s_3,t_3$ such that $s_2
  \StarOf{\rightarrow} s_3$ and $t_2 \rightarrow t_3$ and $R(s_3,t_3)$.
\end{theorem}
\begin{proof}
  By induction on $s_1 \StarOf{\rightarrow} s_2$
  \begin{case}[Base Case]
    Let $R(s_1,t_1)$ and $t_1 \StarOf{\rightarrow} t_2$. As
    $(S,\StarOf{\rightarrow})$ is $R$-bisimilar, there exists $s_2$ such that
    $s_1 \StarOf{\rightarrow} s_2$ and $R(s_2,t_2)$, as desired.
  \end{case}

  \begin{figure}
    \centering
    \includegraphics[scale=.4]{solved-proof.pdf}
    \caption{Diagram showing how we prove the inductive case of $R$-confluence
      of the transitive closure of a rewrite system. $R(s_4,t_4)$ comes from the
      inductive hypothesis. The existance of $s_5$ is guaranteed through
      $=$-confluence. $R(s_5,t_5)$ comes from $R$-bisimilarity.}
    \label{fig:solved-proof}
  \end{figure}

  \begin{case}[Inductive Case]
    Let $R(s_1,t_1)$ and $s_1 \StarOf{\rightarrow} s_2$ and $s_2 \rightarrow
    s_3$ and $t_1 \StarOf{\rightarrow} t_2$. By the induction hypothesis, there
    exists $s_4,t_4$ such that $R(s_4,t_4)$ and $s_3 \StarOf{\rightarrow} s_4$
    and $t_3 \rightarrow t_4$.

    As $(S,\rightarrow)$ is $=$-confluent, we know from prior work
    $(S,\StarOf{\rightarrow})$ is $=$-confluent. Because of this, there exists
    $s_5$ such that $s_3 \StarOf{\rightarrow} s_5$ and $s_4 \StarOf{\rightarrow}
    s_5$. As $(S,\StarOf{\rightarrow})$ is $R$-bisimilar, there exists $t_5$
    such that $t_4 \StarOf{\rightarrow} t_5$ and $R(s_5,t_5)$, as desired. This
    case is diagrammed in Figure~\ref{fig:solved-proof}.
  \end{case}
\end{proof}

@@

The full proof of completeness is contained in the appendix of the full version
of the original optician paper~\cite{extended-version}. The original proof of
$R$-confluence for the transitive closure of $\rightarrow$ previously required
additional assumptions. These unnecessary assumptions have been identified and
removed in this version of the proof. Future work used the proof of completeness
over our lens language to show that quotient bijective lenses are also
synthesizeable~\cite{maina+:quotient-synthesis}. Lastly, while synthesizeability
was not proven for symmetric lenses~\cite{soptician}, such a proof would likely
have proven $R$-confluence in a similar manner. Our proof used approaches
inspired by the original Church-Rosser paper~\cite{Church1936SomePO}.

This work continues a trend in making programming easier through
synthesis~\cite{gulwani2017program}. While synthesis is one approach to make
bidirectional programming easier, it is not the only approach. Recent work has
gone into building lenses without requiring the point-free combinator
style~\cite{10.1007/978-3-319-89884-1_2}. Other work has found
applicative~\cite{10.1145/2858949.2784750} and
monadic~\cite{10.1007/978-3-030-17184-1_6} approaches to compositionally
building lenses.

>>>>

  The proof that all string lenses are synthesizeable involves proving a
  confluence-like property, parameterized by a an additional binary relation
  $R$. We call this property $R$-confluence. In this model, standard confluence
  is the specific case where $R$ is equality. In this paper, we show how
  existing techniques for demonstrating confluence do not work in the domain of
  $R$-confluence, and find that if the rewrite system is $=$-confluent and
  satisfies a commutativity property with $R$, then the system is $R$-confluent.
\end{abstract}

@@

Optician~\cite{optician}, an extension of Boomerang, makes bidirectional
programming easier by supporting synthesis of bidirectional string
transformations. More specifically, it takes as input two regular expressions
($\Regex$ and $\RegexAlt$, which serve as the type of a Boomerang lens) and a
set of examples specifying input-output behavior, and synthesizes a well-typed
lens between the languages of those regular expressions. For brevity, we will
not provide formal definitions for some aspects of Optician; the interested
reader can find such definitions in the original Optician paper~\cite{optician}.
Furthermore, examples detailing the use of synthesized lenses in practice (which
we also elide for space) can be found in that paper and follow-up
work~\cite{maina+:quotient-synthesis,soptician}.

To indicate a lens $\Lens$ is well typed, and converts between the languages of
$\Regex$ and $\RegexAlt$, we write $\Lens : \Regex \Leftrightarrow \RegexAlt$.
For synthesis, given $\Regex$ and $\RegexAlt$, we must find a lens $\Lens$ such
that $\Lens : \Regex \Leftrightarrow \RegexAlt$. In the context of lens
synthesis, there are three sorts of lens typing rules: syntax-directed rules,
composition, and type equivalence.

@@

In our search algorithm, we only search through equivalent regular expressions
once, before processing any syntax directed-rules. We formalize this in the
typing of DNF regular expressions by only permitting the application of
type-equivalence once, after all syntactic rules have been applied. This is
enforced by having two typing judgements: one for the ``rewriteless'' type of
the lens (meaning no type-equivalence rules were applied) and one of the
``full'' type of the lens. If $\DNFLens \OfRewritelessType \DNFRegex
\Leftrightarrow \DNFRegexAlt$, then $\DNFLens$ is a DNF lens of rewriteless type
$\DNFRegex \Leftrightarrow \DNFRegexAlt$. If $\DNFLens \OfType \DNFRegex
\Leftrightarrow \DNFRegexAlt$, then $\DNFLens$ is a DNF lens of full type
$\DNFRegex \Leftrightarrow \DNFRegexAlt$. The following rule is used to get the
full type of a DNF lens from the rewriteless type.
\begin{mathpar}
  \inferrule[\RewriteDNFRegexLensRule{}]
  {
    \DNFRegex' \StarOf{\Rewrite} \DNFRegex\\
    \DNFRegexAlt' \StarOf{\Rewrite} \DNFRegexAlt \\
    \DNFLens \OfRewritelessType \MapsBetweenTypeOf{\DNFRegex}{\DNFRegexAlt}
  }
  {
    \DNFLens \OfType \MapsBetweenTypeOf{\DNFRegex'}{\DNFRegexAlt'}
  }
\end{mathpar}
Note that instead of using directionless equivalences,
\RewriteDNFRegexLensRule{} uses directioned rewrites. The relationship between
the two is formalized by the following theorem:
\begin{theorem}
  \label{thm:thm1}
  If $\Regex \SSREquiv \RegexAlt$, then $\ToDNFRegexOf{\Regex}
  \equiv_{\rightarrow} \ToDNFRegexOf{\RegexAlt}$, where $\ToDNFRegexOf{\Regex}$
  and $\ToDNFRegexOf{\RegexAlt}$ are $\Regex$ and $\RegexAlt$ in DNF form
  (respectively), and $\equiv_{\rightarrow}$ is the reflexive, transitive, and
  symmetric closure of $\rightarrow$.
\end{theorem}

Proving that our search procedure can generate any lens reduces to proving DNF
lenses complete with respect to our standard lens language. In particular, we
wish the prove:
\begin{theorem}
  If $\Lens : \Regex \Leftrightarrow \RegexAlt$, then there exists a DNF lens,
  $\DNFLens$, such that $\DNFLens \OfType \ToDNFRegexOf{\Regex} \Leftrightarrow
  \ToDNFRegexOf{\RegexAlt}$ and the semantics of $\Lens$ and $\DNFLens$
  are equivalent.
\end{theorem}

We prove this propety by induction on the structure of the typing derivation.
Particular difficulty lies in the lens equivalence rule. We begin this case
below:

\begin{mathpar}
  \inferrule
  {
    \Lens \OfType \Regex_1 \Leftrightarrow \RegexAlt_1\\
    \Regex_1 \SSREquiv \Regex_2\\
    \RegexAlt_1 \SSREquiv \RegexAlt_2
  }
  {
    \Lens \OfType \Regex_2 \Leftrightarrow \RegexAlt_2
  }
\end{mathpar}

@@

We first show that there exist $\DNFRegex_2$ and $\DNFRegexAlt_2$ such that
$\ToDNFRegexOf{\Regex_2} \StarOf{\rightarrow} \DNFRegex_2$ and
$\ToDNFRegexOf{\DNFRegex_1} \StarOf{\rightarrow} \DNFRegex_2$ and
$\ToDNFRegexOf{\RegexAlt_2} \StarOf{\rightarrow} \DNFRegexAlt_2$ and
$\ToDNFRegexOf{\DNFRegexAlt_1} \StarOf{\rightarrow} \DNFRegexAlt_2$. To do this,
we first prove that $\rightarrow$ is confluent. By
Theorem~\ref{thm:thm1}, we know that $\ToDNFRegexOf{\Regex_1} \equiv_\rightarrow
\ToDNFRegexOf{\Regex_2}$, so confluence implies the existance of such a $\DNFRegex_2$ and $\DNFRegexAlt_2$.

\begin{figure}
  \centering
    \includegraphics[scale=.4]{equiv-case.pdf}
    \caption{Diagram constructing a well-typed DNF lens between
      \ToDNFRegexOf{\Regex_2} and \ToDNFRegexOf{\RegexAlt_2}. Single lined
      arrows indicate rewrites ($\rightarrow$), and triple lines indicate
      equivalence ($\equiv_{\rightarrow}$).}
    \label{fig:equiv-case}
\end{figure}

After this, we have $\DNFLens \OfRewritelessType \DNFRegex_1 \Leftrightarrow
\DNFRegexAlt_1$ and $\DNFRegex_1 \StarOf{\rightarrow} \DNFRegex_2$ and
$\DNFRegexAlt_1 \StarOf{\rightarrow} \DNFRegexAlt_2$. If we can prove a
confluence-like property that would show the existance of some $\DNFLens'
\OfRewritelessType \DNFRegex_3 \Leftrightarrow \DNFRegexAlt_3$, where
$\DNFRegex_2 \StarOf{\rightarrow} \DNFRegex_3$ and $\DNFRegexAlt_2
\StarOf{\rightarrow} \DNFRegexAlt_3$, we would be done. This property is
$R$-confluence for a properly chosen $R$ (which we describe in \S\ref{sec:rcf}).
This case is diagrammed in Figure~\ref{fig:equiv-case}.

\section{$R$-Confluence Formulation}
\label{sec:rcf}
Let $S$ be an underlying set, and $\rightarrow$ and $R$ be
binary relations. The rewrite system $(S,\rightarrow)$ is $R$-confluent, if for
all $s_1,s_2 \in S$, if $R(s_1,s_2)$, $s_1 \StarOf{\rightarrow} s_1'$, and $s_2
\StarOf{\rightarrow} s_2'$, then there exist $s_1''$ and $s_2''$ such that
$R(s_1'',s_2'')$, $s_1' \StarOf{\rightarrow} s_1''$, and $s_2
\StarOf{\rightarrow} s_2''$.

@@

Now that $R$-confluence has been formally defined, we can ask ourselves: ``What
is a good approach to proving $R$-confluence?'' One approach is to prove that
our rewrite system is locally confluent, which is equivalent to $=$-confluence
in a terminating system~\cite{huetconf}. Unfortunately, our rewrites are not
terminating, so this approach does not work.

An approach pioneered by Tait and Martin-L\"{o}f~\cite{barendrecht} still works
in non-terminating systems. This approach uses the \emph{diamond property}: A
rewrite system $(S,\rightarrow)$ satisfies the diamond property if $s_1
\rightarrow s_2$ and $s_1 \rightarrow s_3$ implies that there exists $s_4$ such
that $s_2 \rightarrow s_4$, and $s_2 \rightarrow s_4$. If a rewrite system
satisfies the diamond property, then it is also confluent. Unfortunately, this
approach does not work, as the parameterized version of the diamond property
does not imply $R$-confluence.

@@

\begin{figure}
  \centering
    \includegraphics[scale=.4]{cex.pdf}
    \caption{Rewrite system that satisfies the $R$-diamond property, but is not
      $R$-confluent.}
    \label{fig:cex}
\end{figure}

In the Tait and Martin-L\"{o}f approach to proving $(S,\rightarrow)$ confluent,
one must first prove $(S,\rightarrow)$ satisfies the diamond property. Consider
a parameterized property analogous to the diamond property, the
\emph{$R$-diamond property}: A rewrite system $(S,\rightarrow)$ satisfies the
$R$-diamond property if $s_1 \rightarrow s_2$ and $t_1 \rightarrow t_2$ and
$R(s_1,t_1)$ implies that there exists $s_3,t_3$ such that $s_2 \rightarrow s_3$
and $t_2 \rightarrow t_3$ and $R(s_3,t_3)$.

However, satisfying the $R$-diamond property is not sufficient for
$R$-confluence. Consider the simple rewrite system shown in
Figure~\ref{fig:cex}. In this rewrite system, there are 3 elements, $s_1,s_2$,
and $t$. In this setup, $R(s_1,t)$ and $s_1 \rightarrow s_2$. $R$-confluence
requires some $s_3,t'$ such that $s_2 \StarOf{\rightarrow} s_3$ and $t
\StarOf{\rightarrow} t'$ and $R(s_3,t')$, but no such values exist.

To get around this issue, we require a different set of properties.
\begin{enumerate}
\item $(S,\rightarrow)$ must be $=$-confluent.
\item $R$ must be a bisimilulation relation for $(S,\StarOf{\rightarrow})$. In
  other words if $R(s_1,t_1)$, and $s_1 \StarOf{\rightarrow} s_2$, then there
  exists $t_2$ such that $t_1 \StarOf{\rightarrow} t_2$ and $R(s_2,t_2)$; and if
  $R(s_1,t_1)$, and $t_1 \StarOf{\rightarrow} t_2$, then there exists $s_2$ such
  that $s_1 \StarOf{\rightarrow} s_2$ and $R(s_2,t_2)$.
\end{enumerate}

\begin{theorem}
  \label{lem:right-side}
  Let $(S,\rightarrow)$ be $=$-confluent, and $R$ be a bisimilulation relation
  for $(S,\StarOf{\rightarrow})$. If $R(s_1,t_1)$ and $s_1 \StarOf{\rightarrow}
  s_2$ and $t_1 \StarOf{\rightarrow} t_2$, then there exists $s_3,t_3$ such that
  $s_2 \StarOf{\rightarrow} s_3$ and $t_2 \rightarrow t_3$ and $R(s_3,t_3)$.
\end{theorem}
\begin{proof}
  By induction length of the reduction of $s_1 \StarOf{\rightarrow} s_2$
  \begin{case}[Base Case]
    Let $R(s_1,t_1)$ and $t_1 \StarOf{\rightarrow} t_2$. As
    $R$ is a bisimulation relation for $(S,\StarOf{\rightarrow})$, there exists $s_2$ such that
    $s_1 \StarOf{\rightarrow} s_2$ and $R(s_2,t_2)$, as desired.
  \end{case}

  \begin{figure}
    \centering
    \includegraphics[scale=.4]{solved-proof.pdf}
    \caption{Diagram showing how we prove the inductive case of $R$-confluence
      of the transitive closure of a rewrite system. $R(s_4,t_4)$ comes from the
      inductive hypothesis. The existance of $s_5$ is guaranteed through
      $=$-confluence. $R(s_5,t_5)$ is guaranteed because $R$ is a bisimulation
      relation.}
    \label{fig:solved-proof}
  \end{figure}

  \begin{case}[Inductive Case]
    Let $R(s_1,t_1)$ and $s_1 \StarOf{\rightarrow} s_2$ and $s_2 \rightarrow
    s_3$ and $t_1 \StarOf{\rightarrow} t_2$. By the induction hypothesis, there
    exists $s_4,t_4$ such that $R(s_4,t_4)$ and $s_2 \StarOf{\rightarrow} s_4$
    and $t_2 \rightarrow t_4$.

    Because $(S,\rightarrow)$ is $=$-confluent, there
    exists $s_5$ such that $s_3 \StarOf{\rightarrow} s_5$ and $s_4
    \StarOf{\rightarrow} s_5$. As $(S,\StarOf{\rightarrow})$ is a bisimilulation
    relation on $(S,\StarOf{\rightarrow})$,
    there exists $t_5$ such that $t_4 \StarOf{\rightarrow} t_5$ and
    $R(s_5,t_5)$, as desired. This case is diagrammed in
    Figure~\ref{fig:solved-proof}.
  \end{case}
\end{proof}

@@

The concept of $R$-confluence is related to the notion of \emph{confluence
  modulo $\sim$}~\cite{huetconf}. The definition of confluent modulo $\sim{}$ is
almost the same as $\sim$-confluence, the only difference is that confluence
modulo $\sim{}$ requires $\sim{}$ to an equivalence relation. Conditions that suffice to prove a rewrite system confluent modulo
$\sim$ are not generally sufficient to prove $R$-confluence (and vice versa).
Furthermore, our bisimulation relations are closely related to local coherence
modulo $\sim{}$.

Bisimulation relations come from concurrency theory~\cite{bisimcoind}, but a
related notion, commuting rewrites~\cite{commut}, appear in the confluence
literature. We require a single rewrite of $R$ to commute with an arbitrary
number of rewrites of $\rightarrow$, which commuting rewrites do not express.

The full proof of completeness is contained in the appendix of the full version
of the original optician paper~\cite{extended-version}. The original proof of
$R$-confluence for the transitive closure of $\rightarrow$ required
additional assumptions. These unnecessary assumptions have been identified and
removed in this paper. Future work used the proof of completeness
over our lens language to show that quotient bijective lenses are also
synthesizeable~\cite{maina+:quotient-synthesis}. Lastly, while synthesizeability
was not proven for symmetric lenses~\cite{soptician}, such a proof would likely
have proven $R$-confluence in a similar manner.

This work continues a trend in making programming easier through
synthesis~\cite{gulwani2017program}. While synthesis is one approach to make
bidirectional programming easier, it is not the only approach. Work has
gone into building lenses without requiring a point-free combinator
style~\cite{10.1007/978-3-319-89884-1_2}. Other work has found
applicative~\cite{10.1145/2858949.2784750} and
monadic~\cite{10.1007/978-3-030-17184-1_6} approaches to compositionally
building lenses.
